Formula translation in C++, Python and modern Fortran: a~case~study~of~the~language~choice tradeoffs

### 2013-07-15 paper/doc/paper.tex
the new order

%%%%% <gmd>
\documentclass[gmd]{copernicus}%_discussions}
%%%%% </gmd>

@@

% for source code listings
\usepackage{fancyvrb}
\fvset{
  frame=single,
  fontsize=\relscale{.666},
  fontfamily=courier,
  framerule=.2mm,
  framesep=.7mm,
  numbers=left,
  xleftmargin=4mm,
  numbersep=.25mm,
  firstnumber=last
}
\input{pygments}

@@

\newcommand{\codecpp}[4]{%
  \addtocounter{lstnocpp}{1}%
  \renewcommand*\FancyVerbStartString{\PY{c+c1}{//#2}}%
  \renewcommand*\FancyVerbStopString{\PY{c+c1}{//#3}}%
  \setcounter{FancyVerbLine}{\thelinenocpp}%
  \fvset{label={listing~C.\thelstnocpp~(C++)},rulecolor=\color{black},stepnumber=#4}%
  \input{#1}%
  \setcounter{linenocpp}{\value{FancyVerbLine}}%
}
\newcommand{\codepyt}[4]{%
  \addtocounter{lstnopyt}{1}%
  \renewcommand*\FancyVerbStartString{\PY{c}{\PYZsh{}#2}}
  \renewcommand*\FancyVerbStopString{\PY{c}{\PYZsh{}#3}}
  \setcounter{FancyVerbLine}{\thelinenopyt}%
  \fvset{label={listing~P.\thelstnopyt~(Python)},rulecolor=\color{blue},stepnumber=#4}%
  \input{#1}%
  \setcounter{linenopyt}{\value{FancyVerbLine}}%
}
\newcommand{\codefor}[4]{%
  \addtocounter{lstnofor}{1}%
  \renewcommand*\FancyVerbStartString{\PY{c}{!#2}}
  \renewcommand*\FancyVerbStopString{\PY{c}{!#3}}
  \setcounter{FancyVerbLine}{\thelinenofor}%
  \fvset{label={listing~F.\thelstnofor~(Fortran)},rulecolor=\color{red},stepnumber=#4}%
  \input{#1}%
  \setcounter{linenofor}{\value{FancyVerbLine}}%
}

@@

\journal{Computer Physics Communications}

@@

    \author[1]{Sylwester Arabas}
    \author[1]{Dorota Jarecka}
    \author[1]{Anna Jaruga}
    \author[2]{Maciej Fijałkowski}

@@

    \correspondence{Sylwester Arabas (sarabas@igf.fuw.edu.pl)}
    \runningauthor{Arabas, Jarecka et al.}
    \runningtitle{Formula translation in C++, Python \& Fortran}

    \maketitle

    \begin{abstract}
        Three object-oriented implementations of a prototype solver of the advection equation are introduced.
        The presented programs are based on Blitz++ (C++), NumPy (Python), and Fortran's built-in array containers.
        The solvers constitute implementations of the Multidimensional Positive-Definite 
          Advective Transport Algorithm (MPDATA).
        The introduced codes serve as examples for how the application of 
          object-oriented programming (OOP) techniques allows to reproduce the mathematical notation 
          used in the literature within the program code.
        A discussion on the tradeoffs of the programming language choice is presented.
        The main angles of comparison are code brevity and syntax clarity
          (and hence maintainability and auditability) as well as performance.
        All performance tests are carried out using free and open-source compilers.
        In the case of Python, a significant performance gain is observed when switching from the standard 
          interpreter (CPython) to the PyPy implementation of Python.
        Entire source code of all three implementations is embedded in the text and is licensed
          under the terms of the GNU GPL license. 
    \end{abstract}

%    \begin{keyword}
%      object-oriented programming, advection equation, MPDATA, C++, Fortran, Python
%    \end{keyword}

  \introduction

@@

  Application of OOP techniques in development of numerical modelling software may help to:
  \begin{enumerate}[label=(\roman*), leftmargin=*, widest=ii]
    \item{maintain modularity and separation of program logic layers (e.g. separation of
      numerical algorithms, parallelisation mechanisms, data input/output, error handling and
      the description of physical processes); and}
    \item{{\bf shorten and simplify the source code and improve its readability by reproducing within 
      the program logic the mathematical notation used in the literature}.}
  \end{enumerate}
  The first application is attainable, yet arguably cumbersome, with procedural programming.
  The latter, virtually impossible to obtain with procedural programming, is the focus of this paper.
  The importance of reproducing the mathematical notation in the code lays primarily
    in the fact that code readability and brevity \citep[and not the language choice itself,][]{Seeley_2004}
    significantly contribute to code maintainability \citep{Wilson_et_al_2012}.
  Code maintainability, in turn, plays a vital role in community programming efforts 
    and hence geoscientific model development in particular.

  The key aim of this paper is to show how OOP techniques can be used to faithfully
    reproduce within the code what can be referred to as {\em blackboard abstractions} \citep{Rouson_et_al_2012}.
  For this purpose, a sample implementation of a numerical scheme for solving
    the advection equation is introduced in in C++, Python and modern Fortran --
    OOP languages commonly used in scientific computing \citep[see e.g.][chapt.~8]{Einarsson_2005}. 
  Presented benchmark tests, although quite simplistic, together with the experience gained 
    from the development of codes in three different
    languages provide a basis for discussion on the tradeoffs of programming language choice.
  The discussion concerns in principle the development of finite-difference solvers for 
    partial differential equations, but is likely applicable to some extent to the scientific software in general.

  All three programs include an equally structured implementation of the two-dimensional version of the 
    Multidimensional Positive Definite Advective Transport Algorithm \citep[MPDATA;][]{Smolarkiewicz_1984}.
  MPDATA is an example of a numerical procedure used in weather, climate, ocean and solar simulation systems
    \citep[e.g.][respectively]{Ziemianski_et_al_2011,Abiodun_et_al_2011,Ezer_et_al_2002,Charbonneau_and_Smolarkiewicz_2013}.
  Even in its most basic form presented herein, MPDATA is complex enough to contain
    a wide range of mathematical abstractions that can be represented using OOP constructs, 
    yet it is simple enough to allow inclusion of the entire source code within the paper text.
  All relevant MPDATA formul\ae~are given in the text alongside corresponding code fragments
    allowing comparison of the relevant syntax with the mathematical notation.
  In compliance with the article focus, these formul\ae~are presented without derivation or detailed discussion
    \citep[see][for a recent review of MPDATA-based techniques including an introductory description 
    of the algorithm and an exhaustive list of references]{Smolarkiewicz_2006}.

  The paper is structured as follows.
  In section~\ref{sec:impl} we introduce the three implementations
    briefly describing the algorithm itself and
    discussing where and how the OOP techniques may be applied in its implementation.
  Subsections \ref{sec:array}-\ref{sec:solver} describe all three implementations,
    while subsequent sections \ref{sec:cyclic}-\ref{sec:example} cover discussion of C++ code only.
  The relevant parts of Python and Fortran codes do not differ significantly, and for readability reasons 
    are presented in appendices \ref{app:P} and \ref{app:F}, respectively.
  Section~\ref{sec:perf} of this paper covers performance evaluation of the three implementations.
  Section~\ref{sec:tradeoffs} covers discussion of the tradeoffs of the programming language choice.
  Section~\ref{sec:concl} section closes the article with a brief summary.
  Those readers not interested in technical details may wish to skip to section~\ref{sec:perf} or
    \ref{sec:tradeoffs}.

@@

  \section{Implementation}\label{sec:impl}

@@

  MPDATA is a solver for systems of advection equations of the following form:
  \begin{equation}\label{eq:adv}
    \input{formulae_adv}
  \end{equation}
  that describe evolution of a scalar field $\psi$ transported by the fluid flow with velocity $\vec{v}$.
  Solution of equation (\ref{eq:adv}) using MPDATA implies discretisation onto a grid of the $\psi$ and 
    the Courant number $\vec{C}=\vec{v}\cdot\frac{\Delta t}{\Delta x}$ fields, where $\Delta t$ is the 
    solver timestep and $\Delta x$ is the grid spacing.

@@

  In NumPy in order to prevent copying of array data during slicing one needs to operate on the
    so-called array views.
  Array views are obtained when indexing the arrays with objects of the Python's
    built-it \prog{slice} type (or tuples of such objects in case of multi-dimensional arrays).
  Python forbids overloading of operators of built-in types such as \prog{slices}, 
    and does not define addition/subtraction operators for \prog{slice} and \prog{int} pairs.
  Consequently, a custom logic has to be defined not only for fractional indexing,
    but also for shifting the slices by integer intervals ($i\pm1$).
  It is implemented here by declaring a \prog{shift} class with the adequate operator overloads:
  \codepyt{code-pyt-listings.py}{listing03}{listing04}{1}
    and two instances of it to represent unity and half
    in expressions like \prog{i+one}, \prog{i+hlf}, where \prog{i} is an instance of \prog{slice}
    \footnote{\label{fnt:slice}One could argue that not using an own implementation of a slice-representing class
    in NumPy is a design flaw -- being able to modify behaviour of a hypothetical numpy.slice class 
    through inheritance would allow to implement the same behaviour as obtained in listing P.3 without the need to represent 
    the unity as a separate object}:
  \codepyt{code-pyt-listings.py}{listing04}{listing05}{1}

@@

  \subsection{Halo regions}

  An \prog{ext()} function is defined in all three implementation, in order to simplify 
    coding of array range extensions:
  \codecpp{code-cpp-listings.hpp}{listing06}{listing07}{1}
  \codepyt{code-pyt-listings.py}{listing05}{listing06}{1}
  \codefor{code-for-listings.f}{listing04}{listing05}{1}
  Consequently, a range depicted by $i\pm1/2$ may be expressed in the code as \prog{ext(i, h)}.
  In all three implementations the \prog{ext()} function accept the second
    argument to be an integer or a ''half'' (cf. section \ref{sec:grid}).

  \subsection{Array index permutations}\label{sec:pi}
  Hereinafter, the $\pi_{a,b}^{d}$ symbol is used to denote a  
    cyclic permutation of an order $d$ of a set $\{a,b\}$.
  It is used to generalise the MPDATA formul\ae~into multiple dimensions
    using the following notation:
  \begin{equation*}\label{eq:pi}
    \input{formulae_pi}
  \end{equation*}

  Blitz++ ships with the \prog{RectDomain} class (aliased here as \prog{idx\_t}) 
    for specifying array ranges in multiple dimensions.
  The $\pi$ permutation is implemented in C++ as a function \prog{pi()}
    returning an instance of \prog{idx\_t}.
  In order to ensure compile-time evaluation, the permutation order is passed 
    via the template parameter \prog{d}
    (note the different order of \prog{i} and \prog{j} arguments in the two template specialisations):
  \codecpp{code-cpp-listings.hpp}{listing07}{listing08}{1}

  NumPy uses tuples of slices for addressing multi-dimensional array with a
    single object.
  Therefore, the following definition of function \prog{pi()} suffices to represent $\pi$:
  \codepyt{code-pyt-listings.py}{listing06}{listing07}{1}

  In the Fortran implementation \prog{pi()} returns a pointer to the array elements specified
    by \prog{i} and \prog{j} interpreted as (i,j) or
    (j,i) depending on the value of the argument \prog{d}.
  In addition to \prog{pi()}, a helper \prog{span()} function returning the 
    length of one of the vectors passed as argument is defined:
  \codefor{code-for-listings.f}{listing05}{listing06}{1}
  The \prog{span()} function is used to shorten the declarations of arrays to be returned 
    from functions in the Fortran implementation (see listings F.11 and F.17--F.20).

@@

  The tasks to be handled by a prototype advection equation solver proposed herein are:
  \begin{enumerate}[label=(\roman*), leftmargin=*, widest=iii]
    \item{storing arrays representing the $\psi$ and $\vec{C}$ fields and any required housekeeping data,}
    \item{allocating and deallocating the required memory,}
    \item{providing access to the solver state,}
    \item{performing the integration by invoking the advection-operator and
      boundary-condition handling routines.}
  \end{enumerate}
  In the following C++ definition of the \prog{solver} structure, task (i) is represented with the definition of the structure
    member fields; task (ii) is split between the \prog{solver}'s constructor and the destructors of \prog{arrvec\_t};
    task (iii) is handled by the accessor methods; task (iv) is handled within the \prog{solve} method:
  \codecpp{code-cpp-listings.hpp}{listing08}{listing09}{1}
  The \prog{solver} structure is an abstract definition (containing a pure virtual method) 
     requiring its descendants to implement at least
     the \prog{advop()} method which is expected to fill \prog{psi[n+1]} with an updated (advected)
     values of \prog{psi[n]}.
  The two template parameters \prog{bcx\_t} and \prog{bcy\_t} allow the solver to operate with 
    any kind of boundary condition structures that fulfil the requirements implied by the calls to the 
    methods of \prog{bcx} and \prog{bcy}, respectively.

  Python definition of the \prog{solver} class follows closely the C++ structure definition:
  \codepyt{code-pyt-listings.py}{listing07}{listing08}{1}
  The key difference stems from the fact that, unlike Blitz++, NumPy does not allow an array
    to have arbitrary index base -- in NumPy the first element is always addressed with 0.
  Consequently, while in C++ (and Fortran) the computational domain is chosen to start at (i=0, j=0)
    and hence a part of the halo region to have negative indices, in Python the halo region starts 
    at (0,0)\footnote{The reason to allow the domain to begin at an arbitrary index
    is mainly to ease debugging in case the code would be used in parallel computations
    using domain decomposition where each subdomain could have its own index base corresponding to
    the location within the computational domain}.
  However, since the whole halo logic is hidden within the solver, such details are not exposed to the
    user.
  The \prog{bcx} and \prog{bcy} boundary-condition specifications are passed to the solver through
    constructor-like \prog{\_\_init\_\_()} method as opposed to template parameters in C++.

  The above C++ and Python prototype solvers in principle allow to operate with any boundary condition
    objects that implement methods called from within the solver.
  This requirement is checked at compile-time in the case of C++, and at run-time in the case of Python.
  In order to obtain an analogous behaviour with Fortran, it is required to
    define, prior to definition of a solver type, 
    an abstract type with deferred procedures having abstract interfaces 
    \citep[sic!, see Table 2.1 in][for a summary of approximate correspondence of OOP 
      nomenclature between Fortran and C++]{Rouson_et_al_2012}:
  \codefor{code-for-listings.f}{listing06}{listing07}{1}
  Having defined the abstract type for boundary-condition objects, 
    a definition of a solver class following closely the C++ and Python counterparts may be provided:
  \codefor{code-for-listings.f}{listing07}{listing08}{1}

  \subsection{Periodic boundaries (C++)}\label{sec:cyclic}

  From this point, only C++ implementation is explained in the main text. 
  The Python and Fortran implementations are included in appendices A and B.

  The solver definition described in section~\ref{sec:solver} requires a given
    boundary condition object to implement a \prog{fill\_halos()} method.
  An implementation of periodic boundary conditions in C++ is provided in the following listing:
  \codecpp{code-cpp-listings.hpp}{listing09}{listing10}{1}

  Listings P.8 and F.8 contain the Python and Fortran counterparts to listing C.9.

  \subsection{Donor-cell formul\ae~(C++)}\label{sec:donor}
  MPDATA is an iterative algorithm in which each iteration takes the form of the 
    so-called donor-cell formula (which itself is a first-order advection scheme).

  MPDATA and donor-cell are explicit forward-in-time algorithms -- they allow to predict $\psi^{[n+1]}$ as a
    function of $\psi^{[n]}$ where $n$ and $n+1$ denote two adjacent time levels.
  The donor-cell scheme may be written as \citep[eq. 2 in][]{Smolarkiewicz_1984}:
  \begin{equation}\label{eq:donor}
    \input{formulae_donor}
  \end{equation}
  where $N$ is the number of dimensions, 
    and F is the so-called flux function \citep[][eq.~3]{Smolarkiewicz_1984}:
  \begin{equation}\label{eq:donor:F}
    \input{formulae_donor-F}
  \end{equation}

  \noindent The flux function takes the following form in C++:
  \codecpp{code-cpp-listings.hpp}{listing10}{listing11}{1}

  \noindent 
  Equation~\ref{eq:donor} is split into the terms under the summation 
    (effectively the 1-dimensional donor-cell formula):
  \codecpp{code-cpp-listings.hpp}{listing11}{listing12}{1}

  \noindent and the actual two-dimensional donor-cell formula:

  \codecpp{code-cpp-listings.hpp}{listing12}{listing13}{1}

  Listings P.9-P11 and F.9-F.13 contain the Python and Fortran counterparts to listings C.12-C.15.

  \subsection{Donor-cell solver (C++)}\label{sec:donorcell_solver}

  As mentioned in the previous section, the donor-cell formula
    constitutes an advection scheme, hence we may use it to create
    a \prog{solver\_donorcell} implementation of the abstract \prog{solver} class: 
  \codecpp{code-cpp-listings.hpp}{listing13}{listing14}{1}
  The above definition is given as an example only.
  In the following sections an MPDATA solver of the same structure is defined.

  Listings P.12 and F.14 contain the Python and Fortran counterparts to listing C.16.

  \subsection{MPDATA formul\ae~(C++)}\label{sec:mpdata}

  MPDATA introduces corrective steps to the algorithm defined by equation~\ref{eq:donor} and \ref{eq:donor:F}.
  Each corrective step is a donor-cell step (eq.~\ref{eq:donor}) with the Courant number
    fields corresponding to the MPDATA antidiffusive velocities of the following 
    form \citep[eqs~13,~14~in][]{Smolarkiewicz_1984}:
  \begin{equation}\label{eq:antidiff1}
    \input{formulae_antidiff1}
  \end{equation}
  where $\psi$ and $C$ represent values from the previous iteration and where:
  \begin{equation}\label{eq:antidiff2}
    \input{formulae_antidiff2}
  \end{equation}
  For positive-definite $\psi$, the $A$ and $B$ terms take the 
  following form\footnote{
    Since $\psi\ge0$, $|A|\le1$ and $|B|\le1$. 
    See \citet[Sec. 4.2]{Smolarkiewicz_2006} for description of adaptation of the 
    formul\ae~for advection of fields of variable sign
  }:
  \begin{equation}\label{eq:A}
    \input{formulae_A}
  \end{equation}
  \begin{equation}\label{eq:B}
    \input{formulae_B}
  \end{equation}

  If the denominator in equations \ref{eq:A} or \ref{eq:B} equals zero for a given {\em i} and {\em j}, 
    the corresponding $A_{[i,j]}$ and $B_{[i,j]}$ are set to zero what may be conveniently 
    represented with the \prog{where} construct (available in all three considered languages):
  \codecpp{code-cpp-listings.hpp}{listing14}{listing15}{1}

  The $A$ term defined in equation \ref{eq:A} takes the following form:
  \codecpp{code-cpp-listings.hpp}{listing15}{listing16}{1}

  The $B$ term defined in equation \ref{eq:B} takes the following form:
  \codecpp{code-cpp-listings.hpp}{listing16}{listing17}{1}

  Equation \ref{eq:antidiff2} takes the following form:
  \codecpp{code-cpp-listings.hpp}{listing17}{listing18}{1}

  Equation \ref{eq:antidiff1} take the following form:
  \codecpp{code-cpp-listings.hpp}{listing18}{listing19}{1}

  Listings P.13-P.17 and F.15-F.21 contain the Python and Fortran counterparts to listing C.16-C.22.

  \subsection{MPDATA solver (C++)}\label{sec:mpdatasolver}

>>>>

%%%%% <sciprog>
\documentclass[twocolumn]{article}
\usepackage{lineno,authblk,latexsym,amsmath,float}
\usepackage[colorlinks=true]{hyperref}
\usepackage[hmargin=.55in,vmargin=.75in]{geometry}
\usepackage[numbers]{natbib}
\setlength{\columnsep}{1cm}
%%%%% </sciprog>

@@

% fr compact section headers
\usepackage[compact]{titlesec}
\titlespacing*{\section}{0pt}{*0}{0pt}

% for source code listings
\usepackage{fancyvrb}
\fvset{
  frame=single,
  fontsize=\relscale{.6},
  fontfamily=courier,
  framerule=.2mm,
  framesep=.7mm,
  numbers=left,
  xleftmargin=4mm,
  numbersep=.25mm,
  firstnumber=last
}
\input{pygments}

@@

\newcommand{\codecpp}[4]{%
  \linespread{.8}
  \addtocounter{lstnocpp}{1}%
  \renewcommand*\FancyVerbStartString{\PY{c+c1}{//#2}}%
  \renewcommand*\FancyVerbStopString{\PY{c+c1}{//#3}}%
  \setcounter{FancyVerbLine}{\thelinenocpp}%
  \fvset{label={listing~C.\thelstnocpp~(C++)},rulecolor=\color{black},stepnumber=#4}%
  \input{#1}%
  \setcounter{linenocpp}{\value{FancyVerbLine}}%
  \vspace{-1.5ex}
  \linespread{2}
}
\newcommand{\codepyt}[4]{%
  \linespread{.8}
  \addtocounter{lstnopyt}{1}%
  \renewcommand*\FancyVerbStartString{\PY{c}{\PYZsh{}#2}}
  \renewcommand*\FancyVerbStopString{\PY{c}{\PYZsh{}#3}}
  \setcounter{FancyVerbLine}{\thelinenopyt}%
  \fvset{label={listing~P.\thelstnopyt~(Python)},rulecolor=\color{blue},stepnumber=#4}%
  \input{#1}%
  \setcounter{linenopyt}{\value{FancyVerbLine}}%
  \vspace{-1.5ex}
  \linespread{2}
}
\newcommand{\codefor}[4]{%
  \linespread{.8}
  \addtocounter{lstnofor}{1}%
  \renewcommand*\FancyVerbStartString{\PY{c}{!#2}}
  \renewcommand*\FancyVerbStopString{\PY{c}{!#3}}
  \setcounter{FancyVerbLine}{\thelinenofor}%
  \fvset{label={listing~F.\thelstnofor~(Fortran)},rulecolor=\color{red},stepnumber=#4}%
  \input{#1}%
  \setcounter{linenofor}{\value{FancyVerbLine}}%
  \vspace{-1.5ex}
  \linespread{2}
}

@@

\linespread{2}

@@

    \author[1]{Sylwester Arabas\footnote{Corresponence to: S. Arabas -- sarabas@igf.fuw.edu.pl -- 0048225546827 -- University of Warsaw, Pasteura 7, 02-093 Warszawa, Poland}}
    \author[1]{Dorota Jarecka}
    \author[1]{Anna Jaruga}
    \author[2]{Maciej Fijałkowski}

@@

    \onecolumn
    \maketitle

    \begin{abstract}
        Three object-oriented implementations of a prototype solver of the advection equation are introduced.
        The presented programs are based on Blitz++ (C++), NumPy (Python), and Fortran's built-in array containers.
        The solvers constitute implementations of the Multidimensional Positive-Definite 
          Advective Transport Algorithm (MPDATA).
        The introduced codes serve as examples for how the application of 
          object-oriented programming (OOP) techniques allows to reproduce the mathematical notation 
          used in the literature within the program code.
        A discussion on the tradeoffs of the programming language choice is presented.
        The main angles of comparison are code brevity and syntax clarity
          (and hence maintainability and auditability) as well as performance.
        All performance tests are carried out using free and open-source compilers.
        In the case of Python, a significant performance gain is observed when switching from the standard 
          interpreter (CPython) to the PyPy implementation of Python.
        Entire source code of all three implementations is embedded in the text and is licensed
          under the terms of the GNU GPL license.

        ~\\
        {\bf Keywords:} object-oriented programming, advection equation, MPDATA, C++, Fortran, Python
    \end{abstract}

  \twocolumn

  \tableofcontents

  \section{Introduction}

@@

  Application of OOP techniques in development of numerical modelling software may help to:
  \begin{enumerate}[label=(\roman*), leftmargin=*, widest=ii]
    \item{maintain modularity and separation of program logic layers (e.g. separation of
      numerical algorithms, parallelisation mechanisms, data input/output, error handling and
      the description of physical processes); and}
    \item{{\bf shorten and simplify the source code and improve its readability by reproducing within 
      the program logic the mathematical notation used in the literature}.}
  \end{enumerate}
  The first application is attainable, yet arguably cumbersome, with procedural programming.
  The latter, virtually impossible to obtain with procedural programming, is the focus of this paper.
  The importance of reproducing the mathematical notation in the code lays primarily
    in the fact that code readability and brevity \citep[and not the language choice itself,][]{Seeley_2004}
    significantly contribute to code maintainability \citep{Wilson_et_al_2012}.

  The key aim of this paper is to show how OOP techniques can be used to faithfully
    reproduce within the code what can be referred to as {\em blackboard abstractions} \citep{Rouson_et_al_2012}.
  For this purpose, a sample implementation of a numerical scheme for solving
    the advection equation is introduced in C++, Python and modern Fortran --
    OOP languages commonly used in scientific computing \citep[see e.g.][chapt.~8]{Einarsson_2005}. 
  Presented implementations and the results of benchmark tests
    provide a basis for discussion on the tradeoffs of programming language choice.
  The discussion concerns in principle the development of finite-difference solvers for 
    partial differential equations, but is likely applicable to some extent to the scientific programming in general.

  All three programs include an equally structured implementation of the two-dimensional version of the 
    Multidimensional Positive Definite Advective Transport Algorithm \citep[MPDATA;][]{Smolarkiewicz_1984}.
  MPDATA is an example of a numerical procedure used in weather, climate, ocean and solar simulation systems
    \citep[e.g.][respectively]{Ziemianski_et_al_2011,Abiodun_et_al_2011,Ezer_et_al_2002,Charbonneau_and_Smolarkiewicz_2013}.
  The basic MPDATA scheme presented herein is complex enough to contain
    a wide range of mathematical abstractions that can be represented using OOP constructs, 
    yet it is simple enough to allow inclusion of the entire source code within the paper text.
  All relevant MPDATA formul\ae~are given in the text alongside corresponding code fragments
    allowing comparison of the relevant syntax with the mathematical notation.
  In compliance with the article focus, these formul\ae~are presented without derivation or detailed discussion
    (see~\citep{Smolarkiewicz_2006} for a recent review of MPDATA-based techniques including an introductory description 
    of the algorithm and an exhaustive list of references).

  The paper is structured as follows.
  In section~\ref{sec:impl} we introduce the ''formula translation'' part of the three implementations
    briefly describing the algorithm itself and
    discussing where and how the OOP techniques were applied in its implementation.
  The remaining part of the implementations -- the solver logic -- is
    presented in appendix~\ref{sec:app}.
  Section~\ref{sec:perf} covers performance evaluation of the three implementations.
  Section~\ref{sec:tradeoffs} covers discussion of the tradeoffs of the programming language choice.
  Section~\ref{sec:concl} closes the article with a brief summary.

@@

  \section{Implementation: the formul\ae}\label{sec:impl}

@@

  MPDATA is, in its most basic form presented herein, a solver for systems of advection equations of the following form:

  \vspace{-5ex}
  \begin{equation}\label{eq:adv}
    \input{formulae_adv}
  \end{equation}
  that describe evolution of a scalar field $\psi$ transported by the fluid flow with velocity $\vec{v}$.
  Solution of equation (\ref{eq:adv}) using MPDATA implies discretisation onto a grid of~the~$\psi$~and~the~Courant 
    number $\vec{C}=\vec{v}\cdot\frac{\Delta t}{\Delta x}$ fields, where $\Delta t$ is the 
    solver timestep and $\Delta x$ is the grid spacing.

@@

  In NumPy in order to prevent copying of array data during slicing one needs to operate on the
    so-called array views.
  Array views are obtained when indexing the arrays with objects of the Python's
    built-it \prog{slice} type (or tuples of such objects in case of multi-dimensional arrays).
  Python forbids overloading of operators of built-in types such as \prog{slices}, 
    and does not define addition/subtraction operators for \prog{slice} and \prog{int} pairs.
  Consequently, a custom logic has to be defined not only for fractional indexing,
    but also for shifting the slices by integer intervals ($i\pm1$).
  It is implemented here by declaring a \prog{shift} class with the adequate operator overloads:
  \codepyt{code-pyt-listings.py}{listing03}{listing04}{1}
    and two instances of it to represent unity and half
    in expressions like \prog{i+one}, \prog{i+hlf}, where \prog{i} is an instance of \prog{slice}%
    \footnote{\label{fnt:slice}One could argue that not using an own implementation of a slice-representing class
    in NumPy is a design flaw -- being able to modify behaviour of a hypothetical numpy.slice class 
    through inheritance would allow to implement the same behaviour as obtained in listing P.3 without the need to represent 
    the unity as a separate object}:
  \codepyt{code-pyt-listings.py}{listing04}{listing05}{1}

@@

  \subsection{Array index permutations}\label{sec:pi}
  Hereinafter, the $\pi_{a,b}^{d}$ symbol is used to denote a  
    cyclic permutation of an order $d$ of a set $\{a,b\}$.
  It is used to generalise the MPDATA formul\ae~into multiple dimensions
    using the following notation:

  \vspace{-5ex}
  \begin{equation*}\label{eq:pi}
    \input{formulae_pi}
  \end{equation*}

  Blitz++ ships with the \prog{RectDomain} class (aliased here as \prog{idx\_t}) 
    for specifying array ranges in multiple dimensions.
  The $\pi$ permutation is implemented in C++ as a function \prog{pi()}
    returning an instance of \prog{idx\_t}.
  In order to ensure compile-time evaluation, the permutation order is passed 
    via the template parameter \prog{d}
    (note the different order of \prog{i} and \prog{j} arguments in the two template specialisations):
  \codecpp{code-cpp-listings.hpp}{listing06}{listing07}{1}

  NumPy uses tuples of slices for addressing multi-dimensional array with a
    single object.
  Therefore, the following definition of function \prog{pi()} suffices to represent~$\pi$:
  \codepyt{code-pyt-listings.py}{listing05}{listing06}{1}

  In the Fortran implementation \prog{pi()} returns a pointer to the array elements specified
    by \prog{i} and \prog{j} interpreted as (i,j) or
    (j,i) depending on the value of the argument \prog{d}.
  In addition to \prog{pi()}, a helper \prog{span()} function returning the 
    length of one of the vectors passed as argument is defined:
  \codefor{code-for-listings.f}{listing04}{listing05}{1}
  The \prog{span()} function is used to shorten the declarations of arrays to be returned 
    from functions in the Fortran implementation (see listings F.7 and F.12--F.15).

@@

  \subsection{Donor-cell formul\ae}\label{sec:donor}

  MPDATA is an iterative algorithm in which each iteration takes the form of the 
    so-called donor-cell formula (which itself is a first-order advection scheme).

  MPDATA and donor-cell are explicit forward-in-time algorithms -- they allow to predict $\psi^{[n+1]}$ as a
    function of $\psi^{[n]}$ where $n$ and $n+1$ denote two adjacent time levels.
  The donor-cell scheme may be written as \citep[eq. 2 in][]{Smolarkiewicz_1984}:

  \vspace{-5ex}
  \begin{equation}\label{eq:donor}
    \input{formulae_donor}
  \end{equation}
  where $N$ is the number of dimensions, 
    and F is the so-called flux function \citep[][eq.~3]{Smolarkiewicz_1984}:

  \vspace{-5ex}
  \begin{equation}\label{eq:donor:F}
    \input{formulae_donor-F}
  \end{equation}

  \noindent The flux function takes the following form in C++:
  \codecpp{code-cpp-listings.hpp}{listing07}{listing08}{1}
  \vspace{-3.5ex}
  Equation~\ref{eq:donor} is split into the terms under the summation 
    (effectively the 1-dimensional donor-cell formula):
  \codecpp{code-cpp-listings.hpp}{listing08}{listing09}{1}
  \vspace{-3.5ex}
  and the actual two-dimensional donor-cell formula:
  \codecpp{code-cpp-listings.hpp}{listing09}{listing10}{1}
  \vspace{-3.5ex}
  The same formul\ae~are expressed as follows in Python:
  \codepyt{code-pyt-listings.py}{listing06}{listing07}{1}
  \codepyt{code-pyt-listings.py}{listing07}{listing08}{1}
  \codepyt{code-pyt-listings.py}{listing08}{listing09}{1}
  \vspace{-3.5ex}
  The Fortran counterparts are: 
  \codefor{code-for-listings.f}{listing05}{listing06}{1}
  \codefor{code-for-listings.f}{listing06}{listing07}{1}
  \codefor{code-for-listings.f}{listing07}{listing08}{1}
  \codefor{code-for-listings.f}{listing08}{listing09}{1}
  \codefor{code-for-listings.f}{listing09}{listing10}{1}

  \subsection{MPDATA formul\ae}\label{sec:mpdata}

  MPDATA introduces corrective steps to the algorithm defined by equation~\ref{eq:donor} and \ref{eq:donor:F}.
  Each corrective step is a donor-cell step (eq.~\ref{eq:donor}) with the Courant number
    fields corresponding to the MPDATA antidiffusive velocities of the following 
    form \citep[eqs~13,~14~in][]{Smolarkiewicz_1984}:

  \vspace{-5ex}
  \begin{equation}\label{eq:antidiff1}
    \input{formulae_antidiff1}
  \end{equation}
  where $\psi$ and $C$ represent values from the previous iteration and where:

  \vspace{-5ex}
  \begin{equation}\label{eq:antidiff2}
    \input{formulae_antidiff2}
  \end{equation}
  For positive-definite $\psi$, the $A$ and $B$ terms take the 
  following form\footnote{
    Since $\psi\ge0$, $|A|\le1$ and $|B|\le1$. 
    See \citet[Sec. 4.2]{Smolarkiewicz_2006} for description of adaptation of the 
    formul\ae~for advection of fields of variable sign
  }:

  \vspace{-5ex}
  \begin{equation}\label{eq:A}
    \input{formulae_A}
  \end{equation}
  \begin{equation}\label{eq:B}
    \input{formulae_B}
  \end{equation}

  If the denominator in equations \ref{eq:A} or \ref{eq:B} equals zero for a given {\em i} and {\em j}, 
    the corresponding $A_{[i,j]}$ and $B_{[i,j]}$ are set to zero what may be conveniently 
    represented with the \prog{where} construct in all three considered languages:
  \codecpp{code-cpp-listings.hpp}{listing10}{listing11}{1}
  \codepyt{code-pyt-listings.py}{listing09}{listing10}{1}
  \codefor{code-for-listings.f}{listing10}{listing11}{1}
  \codefor{code-for-listings.f}{listing11}{listing12}{1}

  The $A$ term defined in equation \ref{eq:A} takes the following form:
  \codecpp{code-cpp-listings.hpp}{listing11}{listing12}{1}
  \codepyt{code-pyt-listings.py}{listing10}{listing11}{1}
  \codefor{code-for-listings.f}{listing12}{listing13}{1}

  The $B$ term defined in equation \ref{eq:B} takes the following form:
  \codecpp{code-cpp-listings.hpp}{listing12}{listing13}{1}
  \codepyt{code-pyt-listings.py}{listing11}{listing12}{1}
  \codefor{code-for-listings.f}{listing13}{listing14}{1}

  Equation \ref{eq:antidiff2} takes the following form:
  \codecpp{code-cpp-listings.hpp}{listing13}{listing14}{1}
  \codepyt{code-pyt-listings.py}{listing12}{listing13}{1}
  \codefor{code-for-listings.f}{listing14}{listing15}{1}

  Equation \ref{eq:antidiff1} takes the following form:
  \codecpp{code-cpp-listings.hpp}{listing14}{listing15}{1}
  \codepyt{code-pyt-listings.py}{listing13}{listing14}{1}
  \codefor{code-for-listings.f}{listing15}{listing16}{1}
  \codefor{code-for-listings.f}{listing16}{listing17}{1}
